\sysfull

### 2019-06-28 paper/TOSN/response-letter/responseLetter.tex

\textbf{TOSN-2019-0002:} {\sl Coala: Dynamic Task-Based Intermittent Execution for Energy-Harvesting Devices}

\bigskip

Dear Editor,\\

We would like to thank for handling our submission and providing valuable review results. 
We are also grateful to the reviewers for their substantial effort spent on evaluating our work. \\

With this letter, we address all comments of the reviewers and point to the modified sections of our revised paper. \\

@@

\pagebreak

\begin{responses}

\referee{In section 2.3, authors mention "these models do not rely on capturing an expensive checkpoint, they are usually faster than the checkpoint-based solution [12, 42]". I don't fully agree with this. It depends on the underlying checkpointing mechanism. If it copies complete stack every time during the checkpoint then yes it's expensive. But checkpointing mechanism like incremental checkpointing [2] (also see the paper by Faycal Ait Aouda), where you only update modified variables, is much better in terms of energy consumption (even better than the task-based solution in my opinion). Also, [12] and [42] are not checkpoint-based solutions.}

\response{As indicated by the reviewer, [12, 42] are not checkpoint systems. They are task-based systems and their results show that task-based systems are faster than checkpoint-based ones. Therefore, we restated their claim and provided the supporting references, which are [12, 42].\\
 Coala addresses deficiencies in the \emph{static} task-based systems and provides a more efficient alternative. Therefore, we compared its performance against the state-of-the-art task-based runtime.}

@@

\referee{In section 7.1, authors mention "produce a high number of complete runs". Please mention the exact number iterations. Also please mention the capacitor size for the WIPS 5.1.}

\response{The number of complete runs, which ranges between 4 and 125, and WISP capacitor size are added to Section 7.1.}

\referee{In figure 10, it is not very clear how authors normalized the values for Coala in case of fft for 30cm and 50cm. Also in the text authors state that "This is marked with $\infty$ signs in Fig. 9", its figure 10. }

\response{We normalized these two special cases on 1 second. This information is added to Section 8.2; additionally, the figure reference is corrected. }

@@

\end{responses}

\begin{additionalQuestions}
\referee{Does this paper cite and use appropriate references?: No\\
If not, what important references are missing?: 
Ratchet and Clank were the first to propose using a timer for breaking up a long task so that it can make forward progress during short power cycles.  The authors propose the idea as if they were the first to use it.}

\response{Ratchet and Clank are mentioned and referenced in our original paper. 
In Section 4.2, we explicitly mentioned that the timer-based approach is similar to [72]; [72] references the ``Ratchet'' paper. 
In Section 9 we explained the core difference between our timer-based approach and the approach explained  in ``Ratchet'' paper. Furthermore, In Section 9 and Table 1 we mentioned Ratchet and Clank by names.}
\end{additionalQuestions}

\begin{editor}
\referee{}
\response{We have added a new figure to the paper, i.e. Fig. 5. The figure's sole purpose is to aid in explaining the concept of \textit{task coalescing}.
We hope that this modification further helps in understanding the core contribution of our paper.}
\end{editor}

### 2019-06-28 paper/TOSN/abstract.tex
working on the rebuttal
minor corrections

Energy-neutral Internet of Things requires freeing embedded devices from batteries and powering them from ambient energy. Ambient energy is, however, unpredictable and can only power a device intermittently. Therefore, the paradigm of intermittent execution is to save the program state into non-volatile memory frequently to preserve the execution progress. In task-based intermittent programming the state is saved at task transition. Tasks are fixed at compile-time and agnostic to energy conditions. Thus, the state may be saved either more often than necessary or not often enough for the program to progress and terminate. To address this challenges, we propose \sys, an adaptive and efficient task-based execution model. \sys progresses on a multi-task scale when energy permits and preserves the computation progress on a sub-task scale if necessary. \sys's specialized memory virtualization mechanism ensures that power failures do not leave the program state in non-volatile memory inconsistent. Our evaluation on a real energy-harvesting platform not only shows that \sys reduces run time by up to 54\% as compared to a state-of-the-art system, but also it is able to progress where static systems fail.

>>>>

Energy-neutral Internet of Things requires freeing embedded devices from batteries and powering them from ambient energy. Ambient energy is, however, unpredictable and can only power a device intermittently. Therefore, the paradigm of intermittent execution is to save the program state into non-volatile memory frequently to preserve the execution progress. In task-based intermittent programming, the state is saved at task transition. Tasks are fixed at compile-time and agnostic to energy conditions. Thus, the state may be saved either more often than necessary or not often enough for the program to progress and terminate. To address these challenges, we propose \sys, an adaptive and efficient task-based execution model. \sys progresses on a multi-task scale when energy permits and preserves the computation progress on a sub-task scale if necessary. \sys's specialized memory virtualization mechanism ensures that power failures do not leave the program state in non-volatile memory inconsistent. Our evaluation on a real energy-harvesting platform not only shows that \sys reduces run time by up to 54\% as compared to a state-of-the-art system, but also it is able to progress where static systems fail.

### 2019-06-28 paper/TOSN/background.tex
related work and minor revision
minor corrections

Energy harvesting devices operate using energy extracted from ambient sources, e.g., radio frequency transmissions, light, etc. These devices elide tethered power and batteries, instead collecting energy into a capacitor, operating when sufficient charge accumulates, and turning off for recharge upon depletion of the buffer. There are several energy harvesting battery-less platforms. For instance, computational RFIDs---open-source TI MSP430-based~\cite{wolverine} WISP~\cite{wisp5} (with its variants such as WISPCam~\cite{naderiparizi_rfid_2015}, NFC-WISP~\cite{zhao_rfid_2015} or NeuralWISP~\cite{holleman_biocas_2008}), Moo~\cite{moo}, and commercial ones such as~\cite{medusa_farsens_2017}. Other platforms include ambient backscatter tag~\cite{liu_sigcomm_2013,parks_sigcomm_2014} or battery-less phone~\cite{talla_imwut_2017}. 
%
% \textbf{Target Hardware of \sys.}
\sys is designed for the demands of existing and future \emph{tiny embedded energy-harvesting platforms} based around general purpose, commodity computing components~\cite{wisp,msp430datasheet}. \sys targets a device with a memory system that has fast, byte-addressable volatile and non-volatile memory; in particular, our target platform, WISP~\cite{wisp}, is equipped with a mixture of SRAM and FRAM. \sys leverages hardware support for fast, bulk-copying between memories via DMA~\cite{msp430datasheet}. \sys does not require architectural additions to commodity processors as in~\cite{su_date_2017,hicks_isca_2017,quickrecall,nvp}.

@@

A power failure clears volatile state (e.g., registers and SRAM) while non-volatile memory (e.g., FRAM) persists. Upon a power failure, control flows to a prior point in the execution: by default, to the beginning of {\tt main()}. Early intermittent systems preserved progress by periodically checkpointing volatile execution context to non-volatile memory~\cite{mementos}, sometimes requiring hardware support~\cite{mottola2017harvos,hibernusplusplus,hibernus,idetic,quickrecall}, Table~\ref{table:chechpoint_comparison} summarizes the data that get copied to and from the non-volatile memory in service of memory consistency and progress preservation. Checkpointing volatile state alone does not ensure data consistency when the system can directly manipulate non-volatile memory~\cite{mspcdino}. Precisely, data can get inconsistent when code includes a \emph{write-after-read} (WAR) dependency between operations that manipulate non-volatile memory. Fig.~\ref{fig:w-a-r} illustrates how state can become inconsistent in an intermittent execution using a simple example of an average operation over an array of integers.
The non-volatile variable \texttt{sum} introduces the WAR, being read and written sequentially by the increment operation. If a power failure occurs right before updating the non-volatile index \texttt{i} (Fig.~\ref{fig:war-execution}, Line 6), \texttt{sum} gets erroneously incremented twice consecutively by the same array element (Fig.~\ref{fig:war-execution}, Lines 5 and 8).

>>>>

Energy harvesting devices operate using energy extracted from ambient sources, e.g., radio frequency transmissions, light, etc. These devices elide tethered power and batteries, instead collecting energy into a capacitor, operating when sufficient charge accumulates, and turning off for recharge upon depletion of the buffer. There are several energy harvesting battery-less platforms. For instance, computational RFIDs---open-source TI MSP430-based~\cite{wolverine} WISP~\cite{wisp5} (with its variants such as WISPCam~\cite{naderiparizi_rfid_2015}, NFC-WISP~\cite{zhao_rfid_2015} or NeuralWISP~\cite{holleman_biocas_2008}), Moo~\cite{moo}, and commercial ones such as~\cite{medusa_farsens_2017}. Other platforms include ambient backscatter tag~\cite{liu_sigcomm_2013,parks_sigcomm_2014,majid2019multi} or battery-less phone~\cite{talla_imwut_2017}. 
%
% \textbf{Target Hardware of \sys.}
\sys is designed for the demands of existing and future \emph{tiny embedded energy-harvesting platforms} based around general purpose, commodity computing components~\cite{wisp,msp430datasheet}. \sys targets a device with a memory system that has fast, byte-addressable volatile and non-volatile memory; in particular, our target platform, WISP~\cite{wisp}, is equipped with a mixture of SRAM and FRAM. \sys leverages hardware support for fast, bulk-copying between memories via DMA~\cite{msp430datasheet}. \sys does not require architectural additions to commodity processors as in~\cite{su_date_2017,hicks_isca_2017,quickrecall,nvp}.

@@

A power failure clears volatile state (e.g., registers and SRAM) while non-volatile memory (e.g., FRAM) persists. Upon power failure, control flows to a prior point in the execution: by default, to the beginning of {\tt main()}. Early intermittent systems preserved progress by periodically checkpointing volatile execution context to non-volatile memory~\cite{mementos}, sometimes requiring hardware support~\cite{mottola2017harvos,hibernusplusplus,hibernus,idetic,quickrecall}, Table~\ref{table:chechpoint_comparison} summarizes the data that get copied to and from the non-volatile memory in service of memory consistency and progress preservation. Checkpointing volatile state alone does not ensure data consistency when the system can directly manipulate non-volatile memory~\cite{mspcdino}. Precisely, data can get inconsistent when code includes a \emph{write-after-read} (WAR) dependency between operations that manipulate non-volatile memory. Fig.~\ref{fig:w-a-r} illustrates how a proram state can become inconsistent in an intermittent execution using a simple example of an average operation over an array of integers.
The non-volatile variable \texttt{sum} introduces the WAR, being read and written sequentially by the increment operation. If a power failure occurs right before updating the non-volatile index \texttt{i} (Fig.~\ref{fig:war-execution}, Line 6), \texttt{sum} gets erroneously incremented twice consecutively by the same array element (Fig.~\ref{fig:war-execution}, Lines 5 and 8).

### 2019-06-28 paper/TOSN/evaluation.tex
minor corrections

For each coalescing strategy from Section~\ref{sec:task_adaptation} (EO, EG,
and WEG) and for a baseline without coalescing (NC), we have measured the time
spent on executing (i) useful task code, (ii) task code wasted due to a power failure, and
(iii) commits to non-volatile memory at the end of each (coalesced)
task.
%
The overhead incurred by each coalescing strategy is broken down in
Fig.~\ref{fig:overallOverheadBreakdown}. Without coalescing enabled (NC), the
re-execution penalty is smallest, because the amount of work that can happen
between commits and may have to be re-executed if interrupted is reduced when
work from multiple static tasks is not combined.
%
However, any gain from a reduced re-execution penalty is canceled out by
the increased commit overhead that is incurred at the end of each static
task.
%
Across all benchmarks, all coalescing strategies reduce more commit overhead
than the re-execution overhead they add.
%
This net overhead reduction is greatest in EG and WEG strategies compared
to the EO strategy. We attribute this discrepancy to EO's slow adjustment
of the coalescing target without regard to the energy conditions.
%
In the subsequent experiments, we focus on the better-performing EG and WEG.

>>>>

For each coalescing strategy from Section~\ref{sec:task_adaptation} (EO, EG,
and WEG) and for a baseline without coalescing (NC), we have measured the time
spent on executing (i) useful task code, (ii) task code wasted due to a power failure, and
(iii) commits to non-volatile memory at the end of each (coalesced)
task.
%
The overhead incurred by each coalescing strategy is broken down in
Fig.~\ref{fig:overallOverheadBreakdown}. Without coalescing enabled (NC), the
re-execution penalty is the smallest, because the amount of work that can happen
between commits and may have to be re-executed if interrupted is reduced when
work from multiple static tasks is not combined.
%
However, any gain from a reduced re-execution penalty is canceled out by
the increased commit overhead that is incurred at the end of each static
task.
%
Across all benchmarks, all coalescing strategies reduce more commit overhead
than the re-execution overhead they add.
%
This net overhead reduction is greatest in EG and WEG strategies compared
to the EO strategy. We attribute this discrepancy to EO's slow adjustment
of the coalescing target without regard to the energy conditions.
%
In the subsequent experiments, we focus on the better-performing EG and WEG.

### 2019-06-28 paper/TOSN/intro.tex
minor corrections

Upon power failures, a device loses the volatile state in its registers, stack, SRAM, and retains the state of any non-volatile memory, such as FRAM. While capturing periodic checkpoints~\cite{mementos,quickrecall} and sleep scheduling~\cite{dewdrop,hibernus,hibernusplusplus} help preserve execution progress, failures can leave non-volatile state incorrect, partially updated. These inconsistencies cause intermittent execution to deviate from continuously-powered behavior, leading to an unrecoverable application failure~\cite{dino,edb}. Prior work developed two main approaches to deal with data inconsistency for intermittently-powered devices: (i) \emph{software-based programming and execution models}~\cite{dino,ratchet,chain,alpaca} and (ii) \emph{hardware-based architectural support}~\cite{hicks_isca_2017,idetic,nvp}. Complex hardware architectural changes are expensive to design, verify, and manufacture. New hardware architectures are also inapplicable to existing systems~\cite{hicks_isca_2017,nvp}. Software approaches are simpler and applicable to existing devices today. Therefore, this work focuses on software approaches. In particular we address the key limitation of task-based programming and execution model, that is the {\em inflexibility} and \emph{energy-unawareness} of statically decomposing a program into tasks.
%
\begin{figure}
    \centering
    \includegraphics[width=0.8\columnwidth]{figures/intro-figure-v5.pdf}
    \caption{Task coalescing reduces commit overhead (left), while task splitting enables termination of large tasks (right).
    S: static legacy task-based runtime, C: \sys (this work).}
    % \caption{Task coalescing reduces time and energy overhead by performing fewer commits (denoted as \texttt{C}) than the individual tasks (denoted as \texttt{Tx}) commits (left). On the other hand, task splitting reduces wasted computation and enables termination of big tasks, requiring more energy than the energy buffer can hold (right).}
    \label{fig:coalesce}
\end{figure}

@@

\paragraph{Drawbacks of Static Task Decomposition.}
Task-based programming and execution models require a programmer~\cite{alpaca,chain} or a compiler~\cite{baghsorkhi_cgo_2018} to statically decompose a program into a collection of tasks. \emph{Tasks}, top level functions, can include arbitrary computation that should be executed despite arbitrarily-timed power failures. The programmer (or a compiler) explicitly expresses task-to-task control flow. Fig.~\ref{fig:coalesce} (left) illustrates how a program's tasks execute and shows how task transitions can affect run time. At each transition, the system incurs an overhead to track and atomically commit modifications to the non-volatile memory, to maintain consistency of program state~\cite{chain,alpaca}. The more task transitions a program requires, the more commit overhead is incurred by the system at run-time. A programmer may thus create very large tasks in an attempt to reduce task transitions overhead. However, a large task may require more energy to complete than a device's fixed hardware energy buffer can hold, which may lead to a task \emph{non-termination} problem (Fig.~\ref{fig:coalesce}, right). To eliminate this risk, existing systems require the programmer to decompose a program into small tasks to preserve execution progress. These constraints on task sizing lead to the following \emph{dilemma}: should large tasks be used, that are efficient but risk non-termination, or small tasks that are guaranteed to complete but incur a high task transition and commit overhead?

\paragraph{Challenges and Contributions.}
We introduce \sys: a new task-based system that employs \emph{adaptive task size execution by task coalescing and splitting}. By means of this novel technique, small tasks can be executed efficiently by trimming unnecessary overhead \emph{dynamically}, meanwhile avoiding the risk of non-termination. \sys accepts any static decomposition and it coalesces (groups) tasks or splits them (see again Fig.~\ref{fig:coalesce}) based on the estimated energy availability without demanding any hardware support. To the best of our knowledge, \sys is the only system that eliminates restructuring and re-compilation of applications considering device's energy buffer---enabling energy-storage independency for the applications, while keeping execution efficient.

>>>>

Upon power failures, a device loses the volatile state in its registers, stack, SRAM, and retains the state of any non-volatile memory, such as FRAM. While capturing periodic checkpoints~\cite{mementos,quickrecall} and sleep scheduling~\cite{dewdrop,hibernus,hibernusplusplus} help preserve execution progress, failures can leave non-volatile state incorrect, partially updated. These inconsistencies cause intermittent execution to deviate from continuously-powered behavior, leading to an unrecoverable application failure~\cite{dino,edb}. Prior work developed two main approaches to deal with data inconsistency for intermittently-powered devices: (i) \emph{software-based programming and execution models}~\cite{dino,ratchet,chain,alpaca} and (ii) \emph{hardware-based architectural support}~\cite{hicks_isca_2017,idetic,nvp}. Complex hardware architectural changes are expensive to design, verify, and manufacture. New hardware architectures are also inapplicable to existing systems~\cite{hicks_isca_2017,nvp}. Software approaches are simpler and applicable to existing devices today. Therefore, this work focuses on software approaches. In particular, we address the key limitation of task-based programming and execution model, that is the {\em inflexibility} and \emph{energy-unawareness} of statically decomposing a program into tasks.
%
\begin{figure}
    \centering
    \includegraphics[width=0.8\columnwidth]{figures/intro-figure-v5.pdf}
    \caption{Task coalescing reduces commit overhead (left), while task splitting enables termination of large tasks (right).
    S: static legacy task-based runtime, C: \sys (this work).}
    % \caption{Task coalescing reduces time and energy overhead by performing fewer commits (denoted as \texttt{C}) than the individual tasks (denoted as \texttt{Tx}) commits (left). On the other hand, task splitting reduces wasted computation and enables termination of big tasks, requiring more energy than the energy buffer can hold (right).}
    \label{fig:coalesce}
\end{figure}

@@

\paragraph{Drawbacks of Static Task Decomposition.}
Task-based programming and execution models require a programmer~\cite{alpaca,chain} or a compiler~\cite{baghsorkhi_cgo_2018} to statically decompose a program into a collection of tasks. \emph{Tasks}, top-level functions, can include arbitrary computation that should be executed despite arbitrarily-timed power failures. The programmer (or a compiler) explicitly expresses task-to-task control flow. Fig.~\ref{fig:coalesce} (left) illustrates how a program's tasks execute and shows how task transitions can affect run time. At each transition, the system incurs an overhead to track and atomically commit modifications to the non-volatile memory, to maintain consistency of program state~\cite{chain,alpaca}. The more task transitions a program requires, the more commit overhead is incurred by the system at run-time. A programmer may thus create very large tasks in an attempt to reduce task transitions overhead. However, a large task may require more energy to complete than a device's fixed hardware energy buffer can hold, which may lead to a task \emph{non-termination} problem (Fig.~\ref{fig:coalesce}, right). To eliminate this risk, existing systems require the programmer to decompose a program into small tasks to preserve execution progress. These constraints on task sizing lead to the following \emph{dilemma}: should large tasks be used, that are efficient but risk non-termination, or small tasks that are guaranteed to complete but incur a high task transition and commit overhead?

\paragraph{Challenges and Contributions.}
We introduce \sys: a new task-based system that employs \emph{adaptive task size execution by task coalescing and splitting}. By means of this novel technique, small tasks can be executed efficiently by trimming unnecessary overhead \emph{dynamically}, meanwhile avoiding the risk of non-termination. \sys accepts any static decomposition and it coalesces (groups) tasks or splits them (see again Fig.~\ref{fig:coalesce}) based on the estimated energy availability without demanding any hardware support. To the best of our knowledge, \sys is the only system that eliminates restructuring and re-compilation of applications considering a device's energy buffer---enabling energy-storage independency for the applications, while keeping execution efficient.

### 2019-06-28 paper/TOSN/paging.tex
minor corrections

VMM overview is given in Fig.~\ref{figure:mem-man}.
VMM abstracts the physical address space of the non-volatile memory (FRAM) and divides it into \texttt{\underline{private}} and \texttt{\underline{shadow}} (underlined locations are non-volatile) buffers, and each buffer is divided into pages.
\texttt{\underline{private}} holds the consistent version of pages after each commit (and on a reboot), while \texttt{\underline{shadow}} serves as support during the execution of coalesced tasks.
The VMM prohibits applications from directly accessing these buffers.
Instead, it redirects any request to the \texttt{working} buffer located in the volatile memory (SRAM), which has a lower latency and energy cost to access than FRAM.
It populates the \texttt{working} buffer with the privatized pages from non-volatile memory requested by tasks.
When a coalesced task completes, the VMM ensures that {\em dirty}, i.e., modified, pages are committed to FRAM and visible to subsequent tasks.
If a power failure interrupts a task, all temporary modifications to pages in \texttt{working} (and \texttt{\underline{shadow}}) occurred after the last commit are discarded in order to keep data consistency.
On a commit, the VMM atomically copies, through a two-phase commit, all dirty pages back into their location in the non-volatile memory.
% SRAM has a limited size and there is an extended working memory containing {\em shadow pages} in the larger FRAM. A shadow page stores the data from an updated page that needs to be committed back to non-volatile memory, but that exceeded the size of the volatile working memory.

@@

When accessing a protected variable with \texttt{RP} or \texttt{WP}, the memory manager first searches the variable's page in the \texttt{working} buffer (Fig.~\ref{figure:mem-man}, Step 1).
If the page is not found there, a page fault is incurred and a new page needs to be swapped in.
If the \texttt{working} buffer is full, a page fault on a memory access requires the VMM to swap out one of the pages in the \texttt{working} buffer (a \emph{victim page}) preserving updates made to that page.
If a task modified any byte in the victim page (i.e., using \texttt{WP}), then the page is dirty and its changes have to be persisted to non-volatile memory (Fig.~\ref{figure:mem-man}, Step 2). If the accessed page was previously modified and swapped out since the last power failure, the most recent version of the page is in {\tt \underline{shadow}} (Step 3), otherwise it has to be retrieved from {\tt \underline{private}} (Step 4). Finally, the page is copied to the volatile buffer (Step 5).

>>>>

VMM overview is given in Fig.~\ref{figure:mem-man}.
VMM abstracts the physical address space of the non-volatile memory (FRAM) and divides it into \texttt{\underline{private}} and \texttt{\underline{shadow}} (underlined locations are non-volatile) buffers, and each buffer is divided into pages.
\texttt{\underline{private}} holds the consistent version of pages after each commit (and on a reboot), while \texttt{\underline{shadow}} serves as support during the execution of coalesced tasks.
The VMM prohibits applications from directly accessing these buffers.
Instead, it redirects any request to the \texttt{working} buffer located in the volatile memory (SRAM), which has a lower latency and energy cost to access than FRAM.
It populates the \texttt{working} buffer with the privatized pages from non-volatile memory requested by tasks.
When a coalesced task completes, the VMM ensures that {\em dirty}, i.e., modified, pages are committed to FRAM and visible to subsequent tasks.
If a power failure interrupts a task, all temporary modifications to pages in \texttt{working} (and \texttt{\underline{shadow}}) occurred after the last commit are discarded in order to keep data consistency.
On a commit, the VMM atomically copies, through a two-phase commit, all dirty pages back into their location in the non-volatile memory.

@@

When accessing a protected variable with \texttt{RP} or \texttt{WP}, the memory manager first searches the variable's page in the \texttt{working} buffer (Fig.~\ref{figure:mem-man}, Step 1).
If the page is not found there, a page fault is incurred and a new page needs to be swapped in.
If the \texttt{working} buffer is full, a page fault on memory access requires the VMM to swap out one of the pages in the \texttt{working} buffer (a \emph{victim page}) preserving updates made to that page.
If a task modified any byte in the victim page (i.e., using \texttt{WP}), then the page is dirty and its changes have to be persisted to non-volatile memory (Fig.~\ref{figure:mem-man}, Step 2). If the accessed page was previously modified and swapped out since the last power failure, the most recent version of the page is in {\tt \underline{shadow}} (Step 3), otherwise it has to be retrieved from {\tt \underline{private}} (Step 4). Finally, the page is copied to the volatile buffer (Step 5).
